Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2018-04-18 Paper/main.tex
Moved previous version of paper to paper/old, included new version, removed all /changess{}

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}

\begin{document}

@@

\begin{abstract}
Recently there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon dubbed as \emph{machine bias}, where trained statistical models -- unbeknownst to their creators -- grow to reflect controversial societal asymmetries, such as gender or racial bias. A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying people of color as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of gender bias in AI.

In this paper, we obtained a comprehensive list of job positions from the U.S. Bureau of Labor Statistics and used it to build sentences in the format of \say{He/She is an Engineer} in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, etc. We translate these sentences into English using the Google Translate API, and collect statistics about the frequency of female, male and gender-neutral pronouns in the translated output. We then show that Google Translate exhibits a strong tendency towards male defaults, which is exaggerated for fields suggested to be troubled with gender stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs. We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce the real-world distribution of female workers. In summary, we provide evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.

We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need for augmented current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
\end{abstract}

@@

Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an \say{universal language} \cite{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \cite{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation. The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \cite{gordin2015scientific}. Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Noam Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation. As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential 1966 ALPAC (Automatic Language Processing Advisory Committee) report( \cite{hutchins1986machine}. 
Such research was then disfavoured in the United States, making a re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \cite{koehn2009statistical,Moses2007}. Statistical and example-based machine translation have been on the rise ever since \cite{Bahdanau2014,carl2003recent,Firat2017}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \cite{wu2016google}) amounting to over $200$ million users daily.

In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism. Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data \cite{Chomsky2011,norvig2017chomsky}. Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics \cite{Chomsky2011}:

@@

Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained} \cite{norvig2017chomsky}:

@@

Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society.

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators grow to reflect controversial societal assymetries, is growing into a pressing concern for the modern times, invites us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed. In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration, with the sole purpose of regulating algorithmic discrimination \cite{kirkpatrick2016battling}.

@@

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
	\label{fig:screenshot-gtranslate-hungarian}
\end{figure}

@@

As of 2017, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily \cite{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative \cite{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as \say{sexist} \cite{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes \cite{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique. They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms \cite{hajian2016algorithmic} The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.

@@

@@

% The hungarian punctuation in 'ő' is not rendered in the PDF. This needs to be fixed somehow. The same problem extends to Table 4 (label tab:templates)
We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian \say{ő ápolónő}, where \say{ápolónő} translates to \say{nurse} and \say{ő} is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result \say{she's a nurse} on Google Translate. As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when \say{nurse} is replaced by \say{engineer}. The same basic template can be ported to all other gender neutral languages, as Table \ref{tab:templates} shows. Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work 
% PEDRO: Removed "with as many gender neutral languages as possible"
 with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures \cite{wals} and other sources. Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
(1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}. To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias. This time we resorted to using the Bureau of Labor Statistics' detailed occupations table \cite{BLS2017}, from the United States Department of Labor, for providing such a list. The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones. Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers. We then separated the curated jobs into categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}. Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset. Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.

% Table containing a list of gender neutral languages with information about pronominal gender systems and GTranslate support
\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family 				& Language 				& Pronominal Gender System 	& Supported 	& Tested 		\\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 					& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 					& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 					& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 					& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 					& \checkmark 	& \xmark		\\
									& English 				& \checkmark 				& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 				& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 					& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 					& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 					& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Korean 				& O 						& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Chinese 				& O 						& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 					& \xmark		& \xmark		\\
									& Pipil 				& \xmark 					& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 					& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 				& \checkmark 	& \xmark		\\
									& Ido 					& O 						& \xmark 		& \xmark 		\\ \cline{2-2}
									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \cline{2-2}
									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
	\end{small}
\end{table}

% List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|m{3.5cm}|c|c|c|}
	\hline	
	Category 										& Group 						& \# Occupations 	& Female Participation	\\ \hline \hline
	Education, training, and library 				& Education						& $22$ 				& $73.0\%$				\\ \hline
	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$				\\ \hline
	Office and administrative support				& Service						& $87$ 				& $72.2\%$				\\ \hline
	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$				\\ \hline
	Management 										& Corporate						& $46$ 				& $39.8\%$				\\ \hline
	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 				\\ \hline
	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$				\\ \hline
	Community and social service 					& Service						& $14$ 				& $66.1\%$				\\ \hline
	Sales and related 								& Corporate						& $28$ 				& $49.1\%$				\\ \hline
	Production 										& Production					& $264$ 			& $28.9\%$				\\ \hline
	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$				\\ \hline
	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$				\\ \hline
	Transportation and material moving				& Service						& $70$ 				& $17.3\%$				\\ \hline
	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$				\\ \hline
	Legal											& Legal 						& $7$				& $52.8\%$				\\ \hline
	Protective Service 								& Service 						& $28$ 				& $22.3\%$				\\ \hline
	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$				\\ \hline
	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$				\\ \hline
	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$				\\ \hline
	Personal care and service						& Service 						& $33$ 				& $76.1\%$				\\ \hline
	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 				\\ \hline
	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$				\\ \hline \hline
	Total											& - 							& $1019$			& $41.3\%$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the U.S. Bureau of Labor Statistics \url{https://???}, grouped by category. We obtained a total of $1019$ occupations from $22$ distinct categories. We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
	\label{tab:occupations}
	\end{small}
\end{table}

% List of sentence templates for assessing gender bias in the translation from source language to English
\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Language 	& Sentence template 						\\ \hline \hline
	Malay		& dia adalah $\langle occupation \rangle$ 	\\ \hline
	Estonian	& ta on $\langle occupation \rangle$ 		\\ \hline
	Finnish		& hän on $\langle occupation \rangle$ 		\\ \hline
	Hungarian	& ő $\langle occupation \rangle$ 			\\ \hline
	Armenian	& na $\langle occupation \rangle$ e 		\\ \hline
	Japanese	& $\langle occupation \rangle$ です			\\ \hline
	Turkish		& o bir $\langle occupation \rangle$ 		\\ \hline
	Yoruba		& o jẹ $\langle occupation \rangle$ 		\\ \hline
	Basque		& $\langle occupation \rangle$ da 			\\ \hline
	Swahili		& yeye ni $\langle occupation \rangle$ 		\\ \hline
	Chinese		& ta $\langle occupation \rangle$ 			\\ \hline
	\end{tabular}
	\caption{Templates used to infer gender biases in the translation to the English language.}
	\label{tab:templates}
	}
\end{table}

>>>>

\usepackage{color,soul}
\usepackage{ifthen}
\newboolean{mark}

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
\begin{document}

\newif\ifchanged
\changedtrue
\setboolean{mark}{true}
\DeclareRobustCommand{\changess}[1]{\ifchanged{\hl{#1}}\else{#1 }\fi}
\DeclareRobustCommand{\changess}[1]{\ifthenelse{\boolean{mark}}{\hl{#1}}{#1}}

@@

\begin{abstract}
Recently there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon dubbed as \emph{machine bias}, where trained statistical models -- unbeknownst to their creators -- grow to reflect controversial societal asymmetries, such as gender or racial bias. A significant number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two distinct Asian people and the now infamous case of Google photos' mistakenly classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of gender bias in AI.

In this paper,  we start with a comprehensive list of job positions from the U.S. Bureau of Labor Statistics (BLS) and used it in order to build sentences in constructions like ``He/She is an Engineer'' (where ``Engineer'' is replaced by the job position of interest) in 11 different gender neutral languages such as Hungarian, Chinese, Yoruba, and several others. We translate these sentences into English using the Google Translate API, and collect statistics about the frequency of female, male and gender-neutral pronouns in the translated output. We then show that Google Translate exhibits a strong tendency towards male defaults,  in particular for fields typically associated to unbalanced gender distribution or  stereotypes such as STEM (Science, Technology, Engineering and Mathematics) jobs. We ran these statistics against BLS' data for the frequency of female participation in each job position, in which we show that Google Translate fails to reproduce a real-world distribution of female workers. In summary, we provide experimental evidence that even if one does not expect in principle a 50:50 pronominal gender distribution, Google Translate yields male defaults much more frequently than what would be expected from demographic data alone.

We believe that our study can shed further light on the phenomenon of machine bias and are hopeful that it will ignite a debate about the need to augment current statistical translation tools with debiasing techniques -- which can already be found in the scientific literature.
\end{abstract}

@@

Although the idea of automated translation can in principle be traced back to as long as the 17th century with Ren\'{e} Descartes proposal of an ``universal language'' \citep{dascal1982universal}, machine translation has only existed as a technological field since the 1950s, with a pioneering memorandum by  Warren Weaver \citep{locke1955machine,weaver1955translation} discussing the possibility of employing digital computers to perform automated translation. The now famous Georgetown-IBM experiment followed not long after, providing the first experimental demonstration of the prospects of automating translation by the means of successfully converting more than sixty Russian sentences into English \citep{gordin2015scientific}. Early systems improved upon the results of the Georgetown-IBM experiment by exploiting Noam Chomsky's theory of generative linguistics, and the field experienced a sense of optimism about the prospects of fully automating natural language translation. As is customary with artificial intelligence, the initial optimistic stage was followed by an extended period of strong disillusionment with the field, of which the catalyst was the influential 1966 ALPAC (Automatic Language Processing Advisory Committee) report( \citep{hutchins1986machine}. 
Such research was then disfavoured in the United States, making a re-entrance in the 1970s before the 1980s surge in statistical methods for machine translation \citep{koehn2009statistical,Moses2007}. Statistical and example-based machine translation have been on the rise ever since \citep{Bahdanau2014,carl2003recent,Firat2017}, with highly successful applications such as Google Translate (recently ported to a neural translation technology \citep{wu2016google}) amounting to over $200$ million users daily.

In spite of the recent commercial success of automated translation tools (or perhaps stemming directly from it), machine translation has amounted a significant deal of criticism. Noted philosopher and founding father of generative linguistics Noam Chomsky has argued that the achievements of machine translation, while successes in a particular sense, are \emph{not successes in the sense that science has ever been interested in}: they merely provide effective ways, according to Chomsky, of approximating unanalyzed data \citep{Chomsky2011,norvig2017chomsky}. Chomsky argues that the faith of the MT community in statistical methods is absurd by analogy with a standard scientific field such as physics \citep{Chomsky2011}:

@@

Leading AI researcher and Google's Director of Research Peter Norvig responds to these arguments by suggesting that even standard physical theories such as the Newtonian model of gravitation are, in a sense, \emph{trained} \citep{norvig2017chomsky}:

@@

Chomsky and Norvig's debate \citep{norvig2017chomsky} is a microcosm  of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society.

On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \citep{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \citep{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \citep{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \citep{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \citep{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \citep{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators grow to reflect controversial societal asymmetries, is growing into a pressing concern for the modern times, invites us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed. In the wave of algorithmic bias, some have argued for the creation of some kind of agency in the likes of the Food and Drug Administration, with the sole purpose of regulating algorithmic discrimination \citep{kirkpatrick2016battling}.

@@

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields\citep{WB2014} such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
	\label{fig:screenshot-gtranslate-hungarian}
\end{figure}

@@

As of 2018, Google translate is one of the largest publicly available machine translation tools in existence, amounting 200 million users daily\citep{Gtranslate200daily2017}. Initially relying on United Nations and European Parliament transcripts to gather data, since 2014 Google Translate has inputed content from its users through the Translate Community initiative\citep{TranslateCommunity}. Recently however there has been a growing concern about gender asymmetries in the translation mechanism, with some heralding it as ``sexist'' \citep{AlgorithmGtranslateSexist2018}. This concern has to at least some extent a scientific backup: A recent study has shown that word embeddings are particularly prone to yielding gender stereotypes\citep{bolukbasi2016man}. Fortunately, the researchers propose a relatively simple \emph{debiasing} algorithm with promising results: they were able to cut the proportion of stereotypical analogies from $19\%$ to $6\%$ without any significant compromise in the performance of the word embedding technique. They are not alone: there is a growing effort to systematically discover and resolve issues of algorithmic bias in black-box algorithms\citep{hajian2016algorithmic}. The success of these results suggest that a similar technique could be used to remove gender bias from Google Translate outputs, should it exist. This paper intends to investigate whether it does. We are optimistic that our research endeavors can be used to argue that there is a positive payoff in redesigning modern statistical translation tools.

@@

In this paper we assume that a statistical translation tool should reflect at most the inequality existent in society -- it is only logical that a translation tool will poll from examples that  society produced and, as such, will inevitably retain some of that bias. It has been argued that one's language affects one's knowledge and cognition about the world \citep{kay1984sapir}, and this leads to the discussion that languages that distinguish between female and male genders grammatically may enforce a bias in the person's perception of the world, with some studies corroborating this, as shown in \citep{boroditsky2003sex}, as well some relating this with sexism \citep{thompson2014linguistic} and gender inequalities \citep{santacreu2013female}.

With this in mind, one can argue that a move towards gender neutrality in language and communication should be striven as a means to promote improved gender equality. Thus, in languages where gender neutrality can be achieved -- such as English -- it would be a valid aim to create translation tools that keep the gender-neutrality of texts translated into such a language, instead of defaulting to male or female variants.

We will thus assume throughout this paper that although the distribution of translated gender pronouns may deviate from 50:50, it should not deviate to the extent of misrepresenting the demographics of job positions. That is to say we shall assume that Google Translate incorporates a negative gender bias if the frequency of male defaults overestimates the (possibly unequal) distribution of male employees per female employee in a given occupation.

@@

We shall assume and then show that the phenomenon of gender bias in machine translation can be assessed by mapping sentences constructed in gender neutral languages to English by the means of an automated translation tool. Specifically, we can translate sentences such as the Hungarian ``ő ápolónő'', where ``ápolónő'' translates to ``nurse'' and ``ő'' is a gender-neutral pronoun meaning either he, she or it, to English, yielding in this example the result ``she's a nurse'' on Google Translate. As Figure \ref{fig:screenshot-gtranslate-hungarian} clearly shows, the same template yields a male pronoun when ``nurse'' is replaced by ``engineer''. The same basic template can be ported to all other gender neutral languages, as depicted in Table \ref{tab:templates}. Given the success of Google Translate, which amounts to 200 million users daily, we have chosen to exploit its API to obtain the desired thermometer of gender bias. Also, in order to solidify our results, we have decided to work with a fair amount of gender neutral languages, forming a list of these with help from the World Atlas of Language Structures (WALS) \citep{wals} and other sources. Table \ref{tab:gender-neutral-languages} compiles all languages we chose to use, with additional columns informing whether they
(1) exhibit a pronominal gender system and (2) are supported by Google Translate. Because pronominal gender systems defy the purposes of our technique, such languages have been discarded.

There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}. To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias. Here, we resorted to using the Bureau of Labor Statistics' detailed occupations table \citep{BLS2017}, from the United States Department of Labor, for providing such a list. The values inside, however, had to be expanded since each line contained multiple occupations and sometimes very specific ones. Fortunately this table also provided a percentage of women participation in the jobs shown, for those that had more than 50 thousand workers. We filtered some of these because they were too generic ( ``Computer occupations, all other'', and others) or because they had gender specific words for the profession (``host/hostess'', ``waiter/waitress''). We then separated the curated jobs into broader categories (Artistic, Corporate, Theatre, etc.) as shown in Table \ref{tab:occupations}. Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset. For the occupations that had less than 50 thousand workers, and thus no data about the participation of women, we assumed that its women participation was that of its upper category. Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}. All code and data used to generate and compile the results presented in the following sections will be made publicly available in a Github repository upon the Editors request. They are not linked to external sites to avoid authors identification.

% Table containing a list of gender neutral languages with information about pronominal gender systems and GTranslate support
\begin{table}[H]
\centering
\begin{small}
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family 				& Language 				& Pronominal Gender System 	& Supported 	& Tested 		\\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 					& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 					& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 					& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 					& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 					& \checkmark 	& \xmark		\\
									& English 				& \checkmark 				& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 				& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 					& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 					& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 					& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Korean 				& O 						& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 					& \checkmark 	& \checkmark	\\
									& Chinese 				& O 						& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 					& \xmark		& \xmark		\\
									& Pipil 				& \xmark 					& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 					& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 				& \checkmark 	& \xmark		\\
									& Ido 					& O 						& \xmark 		& \xmark 		\\ \cline{2-2}
									& Lingua Franca Nova 	& \xmark 					& \xmark 		& \xmark 		\\ \cline{2-2}
									& Interlingua 			& \xmark 					& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
    \end{small}
	\caption{Selected gender neutral languages obtained from WALS and other sources.
     Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
\end{table}

% PEDRO: TODO highlight table
% List of occupation categories obtained from the U.S. Bureau of Labor Statistics, with nº of occupations per category and female participation (%) in each category
\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|m{3.5cm}|c|c|c|}
	\hline	
	Category 										& Group 						& \# Occupations 	& Female Participation	\\ \hline \hline
	Education, training, and library 				& Education						& $22$ 				& $73.0\%$				\\ \hline
	Business and financial operations				& Corporate						& $46$ 				& $54.0\%$				\\ \hline
	Office and administrative support				& Service						& $87$ 				& $72.2\%$				\\ \hline
	Healthcare support 								& Healthcare					& $16$ 				& $87.1\%$				\\ \hline
	Management 										& Corporate						& $46$ 				& $39.8\%$				\\ \hline
	Installation, maintenance, and repair 			& Service						& $91$ 				& $4.0\%$ 				\\ \hline
	Healthcare practitioners and technical 			& Healthcare					& $43$ 				& $75.0\%$				\\ \hline
	Community and social service 					& Service						& $14$ 				& $66.1\%$				\\ \hline
	Sales and related 								& Corporate						& $28$ 				& $49.1\%$				\\ \hline
	Production 										& Production					& $264$ 			& $28.9\%$				\\ \hline
	Architecture and engineering 					& STEM 							& $29$ 				& $16.2\%$				\\ \hline
	Life, physical, and social science				& STEM 							& $34$ 				& $47.4\%$				\\ \hline
	Transportation and material moving				& Service						& $70$ 				& $17.3\%$				\\ \hline
	Arts, design, entertainment, sports, and media 	& Arts / Entertainment 			& $37$ 				& $46.9\%$				\\ \hline
	Legal											& Legal 						& $7$				& $52.8\%$				\\ \hline
	Protective Service 								& Service 						& $28$ 				& $22.3\%$				\\ \hline
	Food preparation and serving related 			& Service 						& $17$ 				& $53.8\%$				\\ \hline
	Farming, fishing, and forestry 					& Farming / Fishing / Forestry 	& $13$ 				& $23.4\%$				\\ \hline
	Computer and mathematical						& STEM 							& $16$ 				& $25.5\%$				\\ \hline
	Personal care and service						& Service 						& $33$ 				& $76.1\%$				\\ \hline
	Construction and extraction 					& Construction / Extraction 	& $68$ 				& $3.0\%$ 				\\ \hline
	Building and grounds cleaning and maintenance 	& Service 						& $10$ 				& $40.7\%$				\\ \hline \hline
	Total											& - 							& $1019$			& $41.3\%$				\\ \hline
	\end{tabular}
    \end{small}
	\caption{Selected occupations obtained from the U.S. Bureau of Labor Statistics \url{https://www.bls.gov/cps/cpsaat11.htm}, grouped by category. We obtained a total of $1019$ occupations from $22$ distinct categories. We have further grouped them into broader groups (or \emph{super-categories}) to ease analysis and visualization.}
	\label{tab:occupations}
\end{table}
